---
title: "lasso_model"
author: "Kellianne Ng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load library
```{r}
library(dplyr)
library(tidyr)
library(lubridate)
library(hdm)
library(ggrepel)
library(ggplot2)
```

## Load dataset
```{r}
df = read.csv("../data/final_data.csv")
df = df %>% mutate(caldt = ymd(caldt))
```

```{r}
fit_lasso_model <- function(full_df,
                            target_col = "alpha_ann",
                            fund_col = "crsp_fundno",
                            date_col = "caldt",
                            year_col = "evaluation_year") {
  
  # 1. Basic checks
  required_cols <- c(target_col, fund_col, date_col)
  missing_cols <- setdiff(required_cols, names(full_df))
  
  if (length(missing_cols) > 0) {
    stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
  }
  
  # 2. create evaluation_year
  df <- full_df %>%
    mutate(
      !!date_col := ymd(.data[[date_col]]),
      !!year_col := as.integer(format(.data[[date_col]], "%Y"))
    )
  
  pred_year = max(df[[year_col]], na.rm = TRUE)
  
  # 3. Identify original numeric predictors
  exclude_base <- c(target_col, fund_col, date_col, year_col)
  
  original_predictor_cols <- setdiff(names(df), exclude_base)
  original_predictor_cols <- original_predictor_cols[sapply(df[original_predictor_cols], is.numeric)]
  
  if (length(original_predictor_cols) == 0) {
    stop("No numeric predictor columns found.")
  }
  
  # 4. Create squared terms from original predictors only
  for (col in original_predictor_cols) {
    df[[paste0(col, "_sq")]] <- df[[col]]^2
  }
  
  # 5. Create pairwise interaction terms from original non-squared predictors only
  if (length(original_predictor_cols) >= 2) {
    interaction_pairs <- combn(original_predictor_cols, 2, simplify = FALSE)
    
    for (pair in interaction_pairs) {
      col1 <- pair[1]
      col2 <- pair[2]
      new_name <- paste0(col1, "_x_", col2)
      df[[new_name]] <- df[[col1]] * df[[col2]]
    }
  }
  # 6. Split into training and prediction sets
  train_df <- df %>%
    filter(.data[[year_col]] < pred_year) %>%
    filter(!is.na(.data[[target_col]])) %>%
    arrange(.data[[date_col]])
  
  test_df <- df %>%
    filter(.data[[year_col]] == pred_year) %>%
    arrange(.data[[date_col]])

  # 7. Select predictor columns
  exclude_from_model <- c(target_col, fund_col, date_col, year_col)
  
  predictor_cols <- setdiff(names(train_df), exclude_from_model)
  predictor_cols <- predictor_cols[sapply(train_df[predictor_cols], is.numeric)]
  
  if (length(predictor_cols) == 0) {
    stop("No numeric predictor columns found.")
  }
  
  # 8. Build training and test matrices
  x_train <- train_df %>%
    select(all_of(predictor_cols)) %>%
    mutate(across(everything(), as.numeric))
  
  x_test <- test_df %>%
    select(all_of(predictor_cols)) %>%
    mutate(across(everything(), as.numeric))
  
  y_train <- train_df[[target_col]]
  
  # 9. Fit plug-in Lasso
  lasso_fit <- rlasso(
    x = as.matrix(x_train),
    y = y_train,
    post = FALSE
  )
  
  # 10. Predict for all funds in prediction_year
  predicted_alpha <- as.numeric(
    predict(lasso_fit, newdata = as.matrix(x_test))
  )
  
  # 11. Return required output
  pred_df <- test_df %>%
    transmute(
      evaluation_year = .data[[year_col]],
      crsp_fundno = .data[[fund_col]],
      predicted_alpha = predicted_alpha
    ) %>%
    arrange(desc(predicted_alpha))
  
  return(pred_df)
}
```

# Coefficient & Sparsity Analysis for Lasso 
# Read df
```{r}
df <- read.csv("../data/df_for_shap_analysis.csv")
head(df,3)
```
## Coefficient analysis for Lasso: 
This function for each forecast year 202-2025, 
takes up all the data up to that point
creates evaluation year
creates squared terms and interaction
fits plug in lambda
fits plug in lasso 
extracts the coefficient vector
stores which variables were selected 

Data frame returned:
- returning a dataframe that summarises
1. how many times each variable was selected
2. average absolute coefficient
3. whether squared terms were picked
4. plot the coefficient paths over time 
```{r}
run_lasso_coefficient_analysis <- function(full_df,
                                           start_year = 2020,
                                           end_year = 2025,
                                           target_col = "alpha_ann",
                                           fund_col = "crsp_fundno",
                                           date_col = "caldt",
                                           year_col = "evaluation_year") {
  
  required_cols <- c(target_col, fund_col, date_col)
  missing_cols <- setdiff(required_cols, names(full_df))
  
  if (length(missing_cols) > 0) {
    stop(paste("Missing required columns:", paste(missing_cols, collapse = ", ")))
  }
  
  # 1. Create evaluation_year
  df <- full_df %>%
mutate(
!!date_col := dmy(.data[[date_col]]),
!!year_col := as.integer(format(.data[[date_col]], "%Y"))
)
  
  # 2. Identify original numeric predictors
  exclude_base <- c(target_col, fund_col, date_col, year_col)
  
  original_predictor_cols <- setdiff(names(df), exclude_base)
  original_predictor_cols <- original_predictor_cols[sapply(df[original_predictor_cols], is.numeric)]
  
  if (length(original_predictor_cols) == 0) {
    stop("No numeric predictor columns found.")
  }
  
  # 3. Create squared terms from original predictors only
  for (col in original_predictor_cols) {
    df[[paste0(col, "_sq")]] <- df[[col]]^2
  }
  
  # 4. Create interaction terms from original non-squared predictors only
  if (length(original_predictor_cols) >= 2) {
    interaction_pairs <- combn(original_predictor_cols, 2, simplify = FALSE)
    
    for (pair in interaction_pairs) {
      col1 <- pair[1]
      col2 <- pair[2]
      new_name <- paste0(col1, "_x_", col2)
      df[[new_name]] <- df[[col1]] * df[[col2]]
    }
  }
  
  coef_results <- list()
  
  # 5. Recursive yearly estimation
  for (pred_year in start_year:end_year) {
    
    train_df <- df %>%
      filter(.data[[year_col]] < pred_year) %>%
      filter(!is.na(.data[[target_col]])) %>%
      arrange(.data[[date_col]])
    
    if (nrow(train_df) == 0) next
    
    exclude_from_model <- c(target_col, fund_col, date_col, year_col)
    
    predictor_cols <- setdiff(names(train_df), exclude_from_model)
    predictor_cols <- predictor_cols[sapply(train_df[predictor_cols], is.numeric)]
    
    if (length(predictor_cols) == 0) next
    
    x_train <- train_df %>%
      select(all_of(predictor_cols)) %>%
      mutate(across(everything(), as.numeric))
    
    y_train <- train_df[[target_col]]
    
    # Fit plug-in Lasso
    lasso_fit <- rlasso(
      x = as.matrix(x_train),
      y = y_train,
      post = FALSE
    )
    
    # Extract full coefficient vector
    coef_vec <- coef(lasso_fit)
    
    coef_df <- data.frame(
      evaluation_year = pred_year,
      variable = names(coef_vec),
      coefficient = as.numeric(coef_vec),
      stringsAsFactors = FALSE
    ) %>%
      filter(variable != "(Intercept)")
    
    coef_results[[as.character(pred_year)]] <- coef_df
  }
  
  final_coef_df <- bind_rows(coef_results)
  
  return(final_coef_df)
}
```

```{r}
coef_df <- run_lasso_coefficient_analysis(
  full_df = df,
  start_year = 2020,
  end_year = 2025,
  target_col = "alpha_ann",
  fund_col = "crsp_fundno",
  date_col = "caldt",
  year_col = "evaluation_year"
)

coef_df
```

## Selection summary 
- how often each variable was selected
- whether its the original, squared or interaction term 
- average coefficient magnitude 
```{r}
selection_summary <- coef_df %>%
  mutate(
    selected = coefficient != 0,
    term_type = case_when(
      grepl("_x_", variable) ~ "interaction",
      grepl("_sq$", variable) ~ "squared",
      TRUE ~ "original"
    )
  ) %>%
  group_by(variable, term_type) %>%
  summarise(
    times_selected = sum(selected),
    mean_coefficient = mean(coefficient, na.rm = TRUE),
    mean_abs_coefficient = mean(abs(coefficient), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(times_selected), desc(mean_abs_coefficient))

selection_summary
```

## Sparsity analysis 
## Analysis of how sparse the Lasso model is 
## Out of all candidate predictors, how many are selected each year? 
## What is the proportion of predictors shrunk to zero? 
```{r}
sparsity_summary <- coef_df %>%
  mutate(selected = coefficient != 0) %>%
  group_by(evaluation_year) %>%
  summarise(
    total_predictors = n(),
    selected_predictors = sum(selected),
    zero_predictors = sum(!selected),
    sparsity_rate = mean(!selected),
    selection_rate = mean(selected),
    .groups = "drop"
  )

sparsity_summary

```
## Sparsity analysis by type 
## Which type of variables are often selected (squared, interaction terms, original variables)
```{r}
sparsity_by_type <- coef_df %>%
  mutate(
    selected = coefficient != 0,
    term_type = case_when(
      grepl("_x_", variable) ~ "interaction",
      grepl("_sq$", variable) ~ "squared",
      TRUE ~ "original"
    )
  ) %>%
  group_by(evaluation_year, term_type) %>%
  summarise(
    total_terms = n(),
    selected_terms = sum(selected),
    selection_rate = mean(selected),
    .groups = "drop"
  )

sparsity_by_type

```

```{r}
library(ggplot2)
library(dplyr)
library(scales)

plot_df <- sparsity_by_type %>%
  group_by(evaluation_year) %>%
  mutate(
    selected_share = selected_terms / sum(selected_terms),
    label_share = paste0(round(selected_share * 100, 1), "%")
  ) %>%
  ungroup()

ggplot(plot_df, aes(x = factor(evaluation_year), y = selected_terms, fill = term_type)) +
  geom_col(position = "fill", width = 0.7) +
  geom_text(
    aes(label = label_share),
    position = position_fill(vjust = 0.5),
    size = 4,
    color = "white"
  ) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Share of Selected LASSO Predictors by Term Type",
    x = "Forecast Year",
    y = "Share of Selected Predictors",
    fill = "Term Type"
  ) +
  theme_minimal()

```


## Plot coefficient paths
```{r}
plot_df <- coef_df %>%
  group_by(variable) %>%
  filter(any(coefficient != 0)) %>%
  ungroup()

label_df <- plot_df %>%
  group_by(variable) %>%
  slice_max(order_by = evaluation_year, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(plot_df, aes(x = evaluation_year, y = coefficient, color = variable)) +
  geom_line() +
  geom_point(data = label_df, size = 1.2) +
  geom_label_repel(
    data = label_df,
    aes(label = variable),
    direction = "y",
    label.size = 0.05,
    min.segment.length = 0,
    size = 3,
    max.overlaps = 20
  ) +
  labs(
    x = "Forecast Year",
    y = "Coefficient",
    title = "Lasso Coefficient Paths by Forecast Year"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


```

## Code to find the top 8/10 variables that were selected the most often
```{r}
selection_summary <- coef_df %>%
  mutate(
    selected = coefficient != 0,
    term_type = case_when(
      grepl("_x_", variable) ~ "interaction",
      grepl("_sq$", variable) ~ "squared",
      TRUE ~ "original"
    )
  ) %>%
  group_by(variable, term_type) %>%
  summarise(
    times_selected = sum(selected),
    mean_abs_coefficient = mean(abs(coefficient), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(times_selected), desc(mean_abs_coefficient))

top_vars <- selection_summary %>%
  slice_max(order_by = times_selected, n = 10) %>%
  pull(variable)

```

## Plotting the top 8/ 10 coefficients 
```{r}
plot_df_top <- coef_df %>%
  filter(variable %in% top_vars)

label_df_top <- plot_df_top %>%
  group_by(variable) %>%
  slice_max(order_by = evaluation_year, n = 1, with_ties = FALSE) %>%
  ungroup()

ggplot(plot_df_top, aes(x = evaluation_year, y = coefficient, color = variable)) +
  geom_line() +
  geom_point(data = label_df_top, size = 1.2) +
  geom_label_repel(
    data = label_df_top,
    aes(label = variable),
    direction = "y",
    label.size = 0.05,
    min.segment.length = 0,
    size = 3,
    max.overlaps = 20
  ) +
  labs(
    x = "Forecast Year",
    y = "Coefficient",
    title = "Top Selected Lasso Coefficients"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
Sparsity plot

```{r}

library(ggplot2)
library(scales)

ggplot(sparsity_summary, aes(x = evaluation_year, y = sparsity_rate)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_text(
    aes(label = paste0(round(sparsity_rate * 100, 1), "%")),
    vjust = -0.8,
    size = 4
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0.8, 1.0)
  ) +
  labs(
    title = "LASSO Sparsity Rate by Forecast Year",
    x = "Forecast Year",
    y = "Sparsity Rate"
  ) +
  theme_minimal()
```

