---
title: "OLS and Elastic Net"
author: "Erica"
date: "2026-03-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```   


```{r}
#load library
library(dplyr)
library(tidyr)
library(lubridate)
library(glmnet)
library(ggplot2)
library(lmtest)
```

THINGS TO DO BEFORE U CLICK RUN ALL
1. make sure final_data.csv is in ../data/
2. Make sure u have qqq_returns.csv in ../data/
3. replace model code with ur own code (replace dummy_model and the model name after the function when running the expanding window)

```{r}
#read in data
file_path = "../data/final_data.csv"
df = read.csv(file_path) 
```

## ols function to be copied to replace the code block in eval
```{r}
#example model here
predict_alpha_ols <- function(train_df){
  df <- train_df %>%
    mutate(
      caldt = as.Date(caldt),
      year = year(caldt)
    ) %>%
    arrange(year, crsp_fundno) %>%
    drop_na()

  evaluation_year <- max(df$year)

  train_data <- df %>%
    filter(year < evaluation_year)

  test_data <- df %>%
    filter(year == evaluation_year)

  feature_names <- train_data %>%
    select(-alpha_ann, -crsp_fundno, -caldt, -year) %>%
    colnames()

  formula <- as.formula(
    paste("alpha_ann ~", paste(feature_names, collapse = " + "))
  )

  model <- lm(formula, data = train_data)

  preds <- predict(model, newdata = test_data)

  result <- tibble(
    evaluation_year = evaluation_year,
    crsp_fundno = test_data$crsp_fundno,
    predicted_alpha = preds
  )

  return(result)
}
```

## feature importance of OLS
```{r}
df <- df %>%
    mutate(
      caldt = as.Date(caldt),
      year = year(caldt)
    ) %>%
    arrange(year, crsp_fundno) %>%
    drop_na()

# train-test split
train_df <- df %>% filter (year <= 2024) %>% filter (year >= 2003)
test_df <- df %>% filter (year == 2025)

# define X and y
feature_names <- train_df %>%
  select (-alpha_ann, -crsp_fundno, -caldt, -year) %>%
  colnames()
X_train <- train_df %>% select(all_of(feature_names))
y_train <- train_df$alpha_ann

# fit ols
formula <- as.formula(
  paste("alpha_ann ~", paste(feature_names, collapse = " + "))
)
model <- lm(formula, data = train_df)

# SHAP = beta * (x - mean)
coefs <- coef(model)[-1]
feature_means <- colMeans(X_train)

# compute SHAP on training data
shap_values <- sweep(X_train, 2, feature_means, "-") %>%
  sweep(2, coefs, "*") %>%
  as.data.frame()

# feature importance tibble
abs_shap <- abs(shap_values)
abs_shap_mean <- colMeans(abs_shap) %>%
  sort(decreasing = TRUE)
importance_df <- tibble(
  feature = names(abs_shap_mean),
  importance = abs_shap_mean
)

# plot
ols_featimp_ranking <- 
  ggplot(importance_df, 
         aes(y = reorder(feature, importance),
             x = importance,
             fill = importance)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "firebrick", high = "dodgerblue4") +
  labs(
    title = "OLS SHAP Ranking",
    y = "",
    x = "Mean (|SHAP|)"
  ) +
  theme_minimal()

ols_featimp_ranking

ggsave("../figures/variable_importance_SHAP_OLS.png", width = 16, height = 12)
```

## enet function to be copied to replace the code block in eval
```{r}
#enet
predict_alpha_enet <- function(train_df){
  df <- train_df %>%
    mutate(
      caldt = as.Date(caldt),
      year = year(caldt)
    ) %>%
    arrange(year, crsp_fundno) %>%
    drop_na()

  evaluation_year <- max(df$year)

  train_full <- df %>%
    filter(year < evaluation_year)

  test_data <- df %>%
    filter(year == evaluation_year)

  if(nrow(test_data) == 0) return(NULL)

  feature_names <- train_full %>%
    select(-alpha_ann, -crsp_fundno, -caldt, -year) %>%
    colnames()

  tune_years <- sort(unique(train_full$year))

  split_index <- floor(0.7 * length(tune_years))

  train_years <- tune_years[1:split_index]
  valid_years <- tune_years[(split_index + 1):length(tune_years)]

  train_tune <- train_full %>%
    filter(year %in% train_years)

  valid_tune <- train_full %>%
    filter(year %in% valid_years)

  X_train <- train_tune %>%
    select(all_of(feature_names)) %>%
    as.matrix()

  y_train <- train_tune$alpha_ann

  X_valid <- valid_tune %>%
    select(all_of(feature_names)) %>%
    as.matrix()

  y_valid <- valid_tune$alpha_ann

  alphas <- seq(0, 1, by = 0.1)

  best_mse <- Inf
  best_alpha <- NULL
  best_lambda <- NULL

  for (a in alphas){

    model <- glmnet(
      X_train,
      y_train,
      alpha = a
    )

    lambdas <- model$lambda

    for (l in lambdas){

      preds <- predict(model, newx = X_valid, s = l)
      mse <- mean((y_valid - preds)^2)

      if (mse < best_mse){
        best_mse <- mse
        best_alpha <- a
        best_lambda <- l
      }
    }
  }

  X_full <- train_full %>%
    select(all_of(feature_names)) %>%
    as.matrix()

  y_full <- train_full$alpha_ann

  enet_model <- glmnet(
    X_full,
    y_full,
    alpha = best_alpha,
    lambda = best_lambda
  )

  X_test <- test_data %>%
    select(all_of(feature_names)) %>%
    as.matrix()

  pred_enet <- predict(enet_model, newx = X_test)

  result <- test_data %>%
    transmute(
      evaluation_year = year,
      crsp_fundno = crsp_fundno,
      predicted_alpha = as.numeric(pred_enet)
    )

  param_df <- tibble(
    evaluation_year = evaluation_year,
    alpha = best_alpha,
    lambda = best_lambda,
    mse = best_mse
    )

  return(list(
    predictions = result,
    params = param_df
    ))
}
```

## feature importance for enet
```{r}
run_enet_expanding <- function(df,
                               test_start_year = 2020,
                               test_end_year = 2025){

  df <- df %>%
    mutate(
      caldt = as.Date(caldt),
      year = year(caldt)
    ) %>%
    arrange(caldt) %>%
    drop_na()

  all_predictions_list <- list()
  all_param_list <- list()

  for (yr in test_start_year:test_end_year){

    cat("Processing year:", yr, "\n")
    train_df <- df %>%
      filter(year <= yr)

    if (nrow(train_df) == 0){
      warning(paste("No data for year", yr))
      next
    }
    model_output <- predict_alpha_enet(train_df)

    if (is.null(model_output)) next

    all_predictions_list[[as.character(yr)]] <- model_output$predictions
    all_param_list[[as.character(yr)]] <- model_output$params
  }

  all_predictions <- bind_rows(all_predictions_list) %>%
    arrange(evaluation_year, crsp_fundno)

  all_params <- bind_rows(all_param_list) %>%
    arrange(evaluation_year)

  return(list(
    all_predictions = all_predictions,
    all_params = all_params
  ))
}

results <- run_enet_expanding(df)
all_predictions <- results$all_predictions
all_params <- results$all_params

# select best alpha and lambda
best_global <- all_params %>%
  arrange(mse) %>%
  slice(1)

best_alpha  <- best_global$alpha
best_lambda <- best_global$lambda

# latest model
train_final <- df %>% filter(year <= 2024)

X_final <- as.matrix(train_final[, feature_names])
y_final <- train_final$alpha_ann

final_model <- glmnet(
  X_final,
  y_final,
  alpha = best_alpha,
  lambda = best_lambda
)

# SHAP of final model
# coefficients
coef_vec <- as.matrix(coef(final_model))[-1, , drop = FALSE]
beta <- as.numeric(coef_vec)

# means
mu <- colMeans(X_final)

# SHAP
shap_values <- sweep(X_final, 2, mu, "-") %>%
  sweep(2, beta, "*") %>%
  as.data.frame()

abs_shap <- abs(shap_values)

importance <- colMeans(abs_shap) %>%
  sort(decreasing = TRUE)

imp_df <- tibble(
  feature = names(importance),
  value = importance
)

enet_featimp_ranking <- 
  ggplot(imp_df,
         aes(y = reorder(feature, value),
             x = value,
             fill = value)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "firebrick", high = "dodgerblue4") +
  labs(
    title = "Elastic Net SHAP Ranking",
    y = "",
    x = "Mean (|SHAP|)"
  ) +
  theme_minimal()

enet_featimp_ranking

ggsave("../figures/variable_importance_SHAP_enet.png", width = 16, height = 12)
```




















