---
title: "RFBoost_Models"
author: "Tiffany"
date: "2026-03-12"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(xgboost)
library(randomForest)
library(fastshap)
library(shapviz)
library(SHAPforxgboost)
```

## Load dataset

```{r}
df = read.csv("../data/final_data.csv")
df = df %>% mutate(caldt = ymd(caldt))
head(df)
```

## Main XGBoost Function
```{r}
predict_alpha_boost <- function(df, target = "alpha_ann") {
  df <- df %>% mutate(caldt = as.Date(caldt)) %>% drop_na()
  features <- setdiff(names(df), c("crsp_fundno", "caldt", "alpha_ann"))
  pred_year = max(year(df$caldt))
  train <- df[year(df$caldt) <  pred_year, ] # take all years before pred year to train
  test  <- df[year(df$caldt) == pred_year, ]
  
  # Hyperparameter tuning
  
  n_years <- sort(unique(year(train$caldt)))
  n_feats <- length(features)
  cutoff  <- n_years[floor(length(n_years) * 0.7)] # first 70% - train, last 30 % - split
  tune_train <- train[year(train$caldt) <= cutoff, ]
  tune_validation <- train[year(train$caldt) > cutoff, ]
  
  dtrain <- xgb.DMatrix(data = as.matrix(tune_train[, features]), label = tune_train[[target]])
  dval <- xgb.DMatrix(data = as.matrix(tune_validation[, features]),   label = tune_validation[[target]])
  
  grid <- expand.grid(
      gamma = c(0, 0.1, 0.2),
      eta = c(0.01, 0.05, 0.1),
      max_depth = c(3, 6, 9),
      subsample = c(0.6, 0.8),  
      colsample_bytree = c(0.6, 0.8), 
      min_child_weight = c(1, 5, 10),
      nrounds = c(50, 100, 200)
  )
  grid$mse <- sapply(seq_len(nrow(grid)), function(i) {
    m <- xgb.train(
      params = list(
        objective = "reg:squarederror",
        eta = as.numeric(grid$eta[i]),
        max_depth = as.integer(grid$max_depth[i]),
        gamma = as.numeric(grid$gamma[i]),
        subsample = as.numeric(grid$subsample[i]),
        colsample_bytree = as.numeric(grid$colsample_bytree[i]),
        min_child_weight = as.integer(grid$min_child_weight[i]),
        seed = 42
      ),
      data = dtrain,
      nrounds = as.integer(grid$nrounds[i]),
      verbose = 0
    )
    preds <- predict(m, dval)
    mean((tune_validation[[target]] - preds)^2)
  })
  best <- grid[which.min(grid$mse), ]
  
  dtrain <- xgb.DMatrix(data = as.matrix(train[, features]), label = train[[target]])
  dtest  <- xgb.DMatrix(data = as.matrix(test[, features]))
  
  model <- xgb.train(
    data = dtrain,
    nrounds = as.integer(best$nrounds),
    verbose = 0,
    params = list (
    gamma = as.numeric(best$gamma),
    eta = as.numeric(best$eta),
    max_depth = as.integer(best$max_depth),
    subsample = as.numeric(best$subsample),
    colsample_bytree = as.numeric(best$colsample_bytree),
    min_child_weight = as.integer(best$min_child_weight),
    objective = "reg:squarederror",
    seed = 42
    )
  )

  preds <- predict(model, dtest)
  mse <- mean((test[[target]] - preds)^2)
  #print best parameters for feature importance function
  cat(sprintf("XGBoost Year %d: best gamma=%.1f, eta=%.2f, max_depth=%d, subsample=%.2f, colsample_bytree=%.2f, min_child_weight=%d, nrounds =%d| MSE:%.4f\n", 
              pred_year, best$gamma, best$eta, best$max_depth,best$subsample, best$colsample_bytree, best$min_child_weight,best$nrounds, mse))
  
  results <- data.frame(
    evaluation_year = pred_year,
    crsp_fundno = test$crsp_fundno,
    predicted_alpha  = preds
  )
  return(results)
}
```

## Predict results for xgboost
```{r}
boost_df <- predict_alpha_boost(df)
```
## View Boost results
```{r}
View(boost_df)
```

# Feature Importance - Random Forest

How does it work?
1. Obtain the model's parameters in the last training window based on the output of the predict_alpha functions
2. Train the model with those same optimised parameters to retrieve the SHAP values
3. Compute SHAP values for random subset of sample size 1500 in training data below (to reduce run time of SHAP functions)
4. Created plots based on derived SHAP values will be automatically saved in figures folder

Plots generated:
1. SHAP ranking bar plot
2. Dependence plots
3. Beeswarm plot
4. Feature Importance (only RF) 

# Feature Importance - XGBoost
```{r}
XGB_feat_importance <- function(df, target = "alpha_ann", sample_size = 1500) {
  
  df <- df %>% mutate(caldt = as.Date(caldt)) %>% drop_na()
  features <- setdiff(names(df), c("crsp_fundno", "caldt", "alpha_ann"))
  n_feats   <- length(features)
  pred_year <- max(year(as.Date(df$caldt)))
  
  train <- df %>% filter(year(as.Date(caldt)) < pred_year)
  dtrain <- xgb.DMatrix(data = as.matrix(train[, features]), label = train[[target]])
  
  params <- list(objective = "reg:squarederror",
                   eta = 0.05,
                   max_depth = 3,
                   gamma = 0.1,
                   subsample = 0.60,
                   colsample_bytree = 0.80,
                   min_child_weight = 1,
                   seed = 42
                 )
  
  Boost_FI_model <- xgb.train(
    params = params, # SUBSTITUTE BEST PARAMS OF LAST WINDOW DURING TRAINING
    data = dtrain, # TO ENSURE EVALUATION IS CONDUCTED ON A MODEL
    nrounds= 200,  # SIMILAR TO THE FINAL ONE BY PREDICT_ALPHA_...FUNCTION
    verbose = 0
  )
  
  # compute evaluation metrics -> verify quality of model on OOS data
  test_data <- df %>% filter(year(as.Date(caldt)) == pred_year)
  dtest  <- xgb.DMatrix(data = as.matrix(test_data[, features]))
  y_pred <- predict(Boost_FI_model, dtest)
  mse <- mean((test_data[[target]] - y_pred)^2)
  cat("MSE :", round(mse, 5), "\n")
  cat("RMSE:", round(sqrt(mse), 5), "\n")
  
  # sample the data because SHAP on a full boost can be very slow
  set.seed(42)
  X_sample <- as.matrix(train[sample(nrow(train), min(nrow(train), sample_size)), features])
  
  shap_values <- shap.values(xgb_model = Boost_FI_model, X_train = X_sample)
  shap_long <- shap.prep(xgb_model = Boost_FI_model, X_train = X_sample)
  
  # 1 - SHAP Ranking
  
  imp_df <- shap_long %>%
    group_by(variable) %>%
    summarise(value = mean(abs(value))) %>%
    arrange(desc(value))

  xgb_featimp_ranking <- ggplot(imp_df,
    aes(y = reorder(variable, value), x = value, fill = value)) + 
    geom_col(show.legend = FALSE) +
    geom_text(aes(label = round(value, 4)), nudge_x = 0.001) +
    scale_fill_gradient(low = "firebrick", high = "dodgerblue4") +
    labs(title = "XGBoost SHAP Ranking", y = "", x = "Mean (|SHAP|)") +
    theme_minimal()
  
  ggsave("../figures/xgb_shap_ranking_bar.png", plot = xgb_featimp_ranking, width = 10, height = 8, dpi = 150)
  print(xgb_featimp_ranking)
  
  return(list(shap_vals = shap_values, shap_long = shap_long))
}

# 2 - Dependence Plot

shap_df = read.csv("../data/df_for_shap_analysis.csv")
shap_results <- XGB_feat_importance(shap_df)
xgb_shap_beeswarm <-shap.plot.summary(shap_results$shap_long)
ggsave(filename = "../figures/xgb_shap_beeswarm.png", plot = xgb_shap_beeswarm, width = 10, height = 8, dpi = 150)

fig_list = lapply(names(shap_results$shap_vals$mean_shap_score)[1:8], 
                  shap.plot.dependence, 
                  data_long = shap_results$shap_long, dilute = 3)
png("../figures/xgb_shap_dep_plot.png", width = 16, height = 12, units = "in", res = 150)
gridExtra::grid.arrange(grobs = fig_list, ncol = 2)
dev.off()
```

## Sanity check
```{r}
print(nrow(boost_df))

duplicates_boost <- boost_df %>%
  group_by(crsp_fundno, evaluation_year) %>%
  summarise(count = n(), .groups = 'drop') %>%
  filter(count > 1)

head(duplicates_boost)
```




